---
title: "Exploratory Data Analysis"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r include=FALSE}
library(tidyverse)
library(lubridate)
library(readxl)
library(knitr)
library(patchwork)
library(ggridges)
library(ggplot2)
library(plotly)
library(leaflet)
library(viridis)
library(scales)
library(sf)
library(dplyr)
```

```{r include=FALSE}

# First, let's check the files in our data directory
files <- list.files("data", pattern = "\\.csv$", full.names = TRUE)

cat("Found", length(files), "files in data directory:\n")
cat(paste(files, collapse = "\n"), "\n")

# Define column types based on Citi Bike data structure
col_types <- cols(
  ride_id = col_character(),
  rideable_type = col_character(),
  started_at = col_datetime(format = ""),
  ended_at = col_datetime(format = ""),
  start_station_name = col_character(),
  start_station_id = col_character(),
  end_station_name = col_character(),
  end_station_id = col_character(),
  start_lat = col_double(),
  start_lng = col_double(),
  end_lat = col_double(),
  end_lng = col_double(),
  member_casual = col_character()
)

# Import and combine all monthly data files
citibike_2023_raw <- files %>% 
  map_dfr(read_csv, 
          col_types = col_types,
          .id = "source_file")

# Clean variable names using janitor
citibike_2023 <- citibike_2023_raw %>%
  janitor::clean_names()

```

```{r include=FALSE}
# Quick look at the data

# First few rows:
head(citibike_2023, 5)

# Last few rows:
tail(citibike_2023, 5)

# Check for missing values
missing_summary <- citibike_2023 %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
  arrange(desc(missing_count))

print(missing_summary)

# Check member distribution
member_summary <- citibike_2023 %>%
  count(member_casual) %>%
  mutate(percentage = n / sum(n) * 100)

print(member_summary)

# Check bike type distribution
bike_summary <- citibike_2023 %>%
  count(rideable_type) %>%
  mutate(percentage = n / sum(n) * 100)

print(bike_summary)
```

```{r include=FALSE}
# Create an initial cleaned dataset

citibike_2023_clean <- citibike_2023 %>%
  mutate(
    # ride length
    ride_length_min = as.numeric(difftime(ended_at, started_at, units = "mins")),
    
    # Date/time components
    start_date = as_date(started_at),
    start_hour = hour(started_at),
    start_month = month(started_at, label = TRUE, abbr = FALSE),
    start_day = wday(started_at, label = TRUE, abbr = FALSE),
    start_year = year(started_at),

    # time of day category
    time_of_day = case_when(
      start_hour >= 5 & start_hour < 12 ~ "Morning",
      start_hour >= 12 & start_hour < 17 ~ "Afternoon",
      start_hour >= 17 & start_hour < 21 ~ "Evening",
      TRUE ~ "Night"
    ),

    # bike type cleaned
    bike_type = case_when(
      rideable_type == "electric_bike" ~ "Electric",
      rideable_type == "classic_bike"  ~ "Classic",
      rideable_type == "docked_bike"   ~ "Docked",
      TRUE ~ "Other"
    ),

    # weekend vs weekday
    day_type = ifelse(start_day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
  ) %>%
  # remove problematic/unrealistic rides
  filter(
    ride_length_min >= 1,
    ride_length_min <= 24 * 60,
    started_at < ended_at,
    !(start_station_id == end_station_id & ride_length_min < 2)
  )

cat("Created citibike_2023_clean with", format(nrow(citibike_2023_clean), big.mark=","), "rows\n")

```

```{r include=FALSE}

# Let's check what we actually filtered
cat("Original dataset size:", format(nrow(citibike_2023), big.mark = ","), "\n")
cat("After 'NYC filtering':", format(nrow(citibike_2023_clean), big.mark = ","), "\n")

# Check longitude distribution to confirm NJ location
lng_summary <- citibike_2023_clean %>%
  summarise(
    min_start_lng = min(start_lng, na.rm = TRUE),
    max_start_lng = max(start_lng, na.rm = TRUE),
    median_start_lng = median(start_lng, na.rm = TRUE),
    pct_west_of_74 = mean(start_lng < -74.00) * 100
  )

cat("Start Longitude Range:\n")
cat(sprintf("  Min: %.4f (Westernmost point)\n", lng_summary$min_start_lng))
cat(sprintf("  Max: %.4f (Easternmost point)\n", lng_summary$max_start_lng))
cat(sprintf("  Median: %.4f\n", lng_summary$median_start_lng))
cat(sprintf("  Percentage west of -74.00: %.1f%% (typical NJ boundary)\n", lng_summary$pct_west_of_74))

```

```{r include=FALSE}
# We created citibike_2023_clean in previous code

# Let's check what variables already exist
cat("\nChecking existing variables in citibike_2023_clean:\n")
cat("ride_length_min exists:", "ride_length_min" %in% names(citibike_2023_clean), "\n")
cat("start_date exists:", "start_date" %in% names(citibike_2023_clean), "\n")
cat("time_of_day exists:", "time_of_day" %in% names(citibike_2023_clean), "\n")
cat("bike_type exists:", "bike_type" %in% names(citibike_2023_clean), "\n")

# If the variables don't exist, add them
if (!"ride_length_min" %in% names(citibike_2023_clean)) {
  cat("\nAdding missing variables to citibike_2023_clean...\n")
  
  citibike_2023_clean <- citibike_2023_clean %>%
    mutate(
      ride_length_min = as.numeric(difftime(ended_at, started_at, units = "mins")),
      
      # Extract temporal features
      start_date = as_date(started_at),
      start_hour = hour(started_at),
      start_month = month(started_at, label = TRUE, abbr = FALSE),
      start_day = wday(started_at, label = TRUE, abbr = FALSE),
      start_year = year(started_at),
      
      # Create useful categories
      time_of_day = case_when(
        start_hour >= 5 & start_hour < 12 ~ "Morning",
        start_hour >= 12 & start_hour < 17 ~ "Afternoon",
        start_hour >= 17 & start_hour < 21 ~ "Evening",
        TRUE ~ "Night"
      ),
      
      bike_type = case_when(
        rideable_type == "electric_bike" ~ "Electric",
        rideable_type == "classic_bike" ~ "Classic",
        rideable_type == "docked_bike" ~ "Docked",
        TRUE ~ "Other"
      ),
      
      day_type = ifelse(start_day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
    ) %>%
    # Remove unrealistic trips (if not already done)
    filter(
      ride_length_min >= 1,
      ride_length_min <= 24 * 60,
      started_at < ended_at,
      !(start_station_id == end_station_id & ride_length_min < 2)
    )
  
  cat("Variables added successfully.\n")
} else {
  cat("\nAll variables already exist in citibike_2023_clean.\n")
}

# Final check
cat(sprintf("Final NJ dataset size: %s rides\n", format(nrow(citibike_2023_clean), big.mark = ",")))
cat("Available variables:\n")
cat(paste(names(citibike_2023_clean), collapse = ", "), "\n")

```

```{r include=FALSE}

# Summary statistics by rider type
summary_stats <- citibike_2023_clean %>%
  group_by(member_casual) %>%
  summarise(
    n_rides = n(),
    avg_duration = mean(ride_length_min, na.rm = TRUE),
    median_duration = median(ride_length_min, na.rm = TRUE),
    sd_duration = sd(ride_length_min, na.rm = TRUE),
    total_hours = sum(ride_length_min, na.rm = TRUE) / 60
  ) %>%
  mutate(
    percentage = n_rides / sum(n_rides) * 100,
    avg_duration = round(avg_duration, 1),
    total_hours = round(total_hours, 0)
  )

kable(summary_stats, digits = 1)

```

```{r include=FALSE}

# Using our existing cleaned NJ dataset and adding NJ-specific variables
citibike_nj_analysis <- citibike_2023_clean %>%
  # Add NJ-specific variables
  mutate(
    # Create consistent member_type variable
    member_type = ifelse(member_casual == "member", "Annual Member", "Casual Rider"),
    member_type = factor(member_type, levels = c("Annual Member", "Casual Rider")),
    
    # Create clean time_of_day with proper ordering
    time_of_day = factor(time_of_day, 
                         levels = c("Morning", "Afternoon", "Evening", "Night")),
    
    # Create clean bike_type with proper ordering
    bike_type = factor(bike_type, 
                       levels = c("Classic", "Electric", "Docked", "Other")),
    
    # Add season variable for later use
    season = case_when(
      start_month %in% c("December", "January", "February") ~ "Winter",
      start_month %in% c("March", "April", "May") ~ "Spring",
      start_month %in% c("June", "July", "August") ~ "Summer",
      start_month %in% c("September", "October", "November") ~ "Fall"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    
    # Add day of week for heatmap
    day_of_week = wday(started_at, label = TRUE, abbr = TRUE),
    
    # Add ride length categories for easier analysis
    ride_length_category = case_when(
      ride_length_min < 10 ~ "Short (<10 min)",
      ride_length_min < 30 ~ "Medium (10-30 min)",
      ride_length_min < 60 ~ "Long (30-60 min)",
      TRUE ~ "Very Long (>60 min)"
    ),
    ride_length_category = factor(ride_length_category,
                                  levels = c("Short (<10 min)", "Medium (10-30 min)", 
                                             "Long (30-60 min)", "Very Long (>60 min)")),
    
    # Add NJ area categories based on longitude
    nj_area = case_when(
      start_lng < -74.03 ~ "Jersey City",
      start_lng >= -74.03 ~ "Hoboken",
      TRUE ~ "Other NJ Area"
    ),
    nj_area = factor(nj_area, levels = c("Hoboken", "Jersey City", "Other NJ Area"))
  )

cat(sprintf("Total NJ City Area rides: %s\n", format(nrow(citibike_nj_analysis), big.mark = ",")))
cat(sprintf("Date range: %s to %s\n", 
            min(citibike_nj_analysis$start_date), 
            max(citibike_nj_analysis$start_date)))
```

```{r include=FALSE}

# Member distribution
member_summary <- citibike_nj_analysis %>%
  count(member_type) %>%
  mutate(percentage = n / sum(n) * 100)

kable(member_summary, digits = 1)

# Bike type distribution
bike_summary <- citibike_nj_analysis %>%
  count(bike_type) %>%
  mutate(percentage = n / sum(n) * 100)

kable(bike_summary, digits = 1)

# NJ area distribution
nj_area_summary <- citibike_nj_analysis %>%
  count(nj_area) %>%
  mutate(percentage = n / sum(n) * 100)

kable(nj_area_summary, digits = 1)

cat(sprintf("  Average ride duration: %.1f minutes\n", mean(citibike_nj_analysis$ride_length_min)))
cat(sprintf("  Median ride duration: %.1f minutes\n", median(citibike_nj_analysis$ride_length_min)))
cat(sprintf("  Most common bike type: %s (%.1f%%)\n", 
            bike_summary$bike_type[1], bike_summary$percentage[1]))
cat(sprintf("  Most rides in: %s (%.1f%%)\n", 
            nj_area_summary$nj_area[1], nj_area_summary$percentage[1]))


```


# Analysis 1: Monthly and seasonal patterns in New Jersey (NJ) City 

```{r echo = FALSE}

# Monthly summary
monthly_summary <- citibike_nj_analysis %>%
  group_by(start_month) %>%
  summarise(
    n_rides = n(),
    avg_duration = mean(ride_length_min),
    total_hours = sum(ride_length_min) / 60,
    .groups = "drop"
  ) %>%
  mutate(
    percentage = n_rides / sum(n_rides) * 100,
    month_num = month(as.Date(paste("2023", start_month, "01"), format = "%Y %B %d"))
  ) %>%
  arrange(month_num)

kable(monthly_summary %>% select(start_month, n_rides, percentage, avg_duration), digits = 1)

# Plot: Total rides per month
p_monthly_total <- monthly_summary %>%
  ggplot(aes(x = start_month, y = n_rides)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            vjust = -0.5, size = 3.5) +
  labs(
    title = "Monthly Ride Volume in NJ City (Jan-Jun 2023)",
    subtitle = "Percentage of total rides shown above bars",
    x = "Month",
    y = "Number of Rides"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_y_continuous(labels = comma)

print(p_monthly_total)
```

The NJ Citi Bike 2023 data shows ridership peaking in July and August with over 100,000 rides each, while winter months have the lowest numbers. Average ride durations are longest in late spring and summer, around 12-13 minutes, and shortest in winter. Overall, bike usage rises in warmer months and falls as temperatures drop.

```{r echo = FALSE}
# Monthly trends by member type
monthly_member <- citibike_nj_analysis %>%
  group_by(start_month, member_type) %>%
  summarise(
    n_rides = n(),
    avg_duration = mean(ride_length_min),
    .groups = "drop"
  ) %>%
  group_by(start_month) %>%
  mutate(
    total = sum(n_rides),
    percentage = n_rides / total * 100
  )

# Plot: Monthly trends by member type
p_monthly_member <- monthly_member %>%
  ggplot(aes(x = start_month, y = n_rides, fill = member_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Monthly Ridership by Member Type in NJ City",
    x = "Month",
    y = "Number of Rides",
    fill = "Member Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  scale_fill_manual(values = c("Annual Member" = "#00BFC4", "Casual Rider" = "#F8766D")) +
  scale_y_continuous(labels = comma)

print(p_monthly_member)
```

The monthly ridership in the NJ city area shows that annual members consistently take more rides than casual riders throughout the year. Both groups see an increase in rides during the warmer months, peaking in July and August, before declining in the colder months. Annual members have a sharper rise and higher overall ridership compared to casual riders.

```{r echo = FALSE}
# Seasonal summary
seasonal_summary <- citibike_nj_analysis %>%
  group_by(season) %>%
  summarise(
    n_rides = n(),
    avg_duration = mean(ride_length_min),
    total_hours = sum(ride_length_min) / 60,
    .groups = "drop"
  ) %>%
  mutate(percentage = n_rides / sum(n_rides) * 100)

kable(seasonal_summary, digits = 1)

# Plot: Total rides by season
p_seasonal_total <- seasonal_summary %>%
  ggplot(aes(x = season, y = n_rides, fill = season)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            vjust = -0.5, size = 4) +
  labs(
    title = "Ride Volume by Season in NJ City Area (Jan-Jun 2023)",
    subtitle = "Spring: Mar-May, Summer: Jun only",
    x = "Season",
    y = "Number of Rides"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "none"
  ) +
  scale_fill_viridis_d(option = "C") +
  scale_y_continuous(labels = comma)

print(p_seasonal_total)

```

The NJ City area sees the highest ride volume in summer, accounting for 32% of rides with an average duration of 12.2 minutes. Fall follows with 27.1% of rides, then spring at 24.2%, both having longer average durations than winter. Winter has the lowest ride count at 16.7%, with the shortest average ride duration of 9.5 minutes. Overall, ride frequency and duration peak in warmer seasons.

# Analysis 2: Temporal patterns in NJ City 

```{r echo = FALSE}
# Hourly patterns
hourly_summary <- citibike_nj_analysis %>%
  group_by(start_hour, day_type) %>%
  summarise(
    n_rides = n(),
    avg_duration = mean(ride_length_min),
    .groups = "drop"
  )

# Peak hours
peak_hours <- hourly_summary %>%
  group_by(day_type) %>%
  slice_max(n_rides, n = 3) %>%
  arrange(day_type, desc(n_rides))

kable(peak_hours, digits = 1)

# Plot: Hourly patterns
p_hourly_patterns <- hourly_summary %>%
  ggplot(aes(x = start_hour, y = n_rides, color = day_type)) +
  geom_line(size = 1.2) +
  geom_point(size = 1.5) +
  labs(
    title = "Hourly Ride Patterns by Day Type in NJ City",
    subtitle = "Commute patterns visible on weekdays",
    x = "Hour of Day",
    y = "Number of Rides",
    color = "Day Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "bottom"
  ) +
  scale_color_manual(values = c("Weekday" = "#1f77b4", "Weekend" = "#ff7f0e")) +
  scale_x_continuous(breaks = seq(0, 23, by = 3)) +
  scale_y_continuous(labels = comma)

print(p_hourly_patterns)
```

Weekday ridership displays distinct peaks, with a rise in ridership in the morning between 8–9 AM and a rise in ridership in the late afternoon between 4-6 PM. In contrast, weekend rides follow a smoother, flatter curve. Overall, weekday rides are substantially higher, reflecting typical workday travel behavior, while weekend activity is steadier and more leisure oriented.

```{r echo = FALSE}
# Day of week analysis
dow_summary <- citibike_nj_analysis %>%
  group_by(day_of_week, day_type) %>%
  summarise(
    n_rides = n(),
    avg_duration = mean(ride_length_min),
    .groups = "drop"
  ) %>%
  mutate(percentage = n_rides / sum(n_rides) * 100)

kable(dow_summary, digits = 1)

# Plot: Day of week patterns
p_dow_patterns <- dow_summary %>%
  ggplot(aes(x = day_of_week, y = n_rides, fill = day_type)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Ride Volume by Day of Week in NJ City",
    x = "Day of Week",
    y = "Number of Rides",
    fill = "Day Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "bottom"
  ) +
  scale_fill_manual(values = c("Weekday" = "#1f77b4", "Weekend" = "#ff7f0e")) +
  scale_y_continuous(labels = comma)

print(p_dow_patterns)
```

Ride volumes steadily rise from Monday and peak midweek on Wednesday and Thursday before decreasing slightly on Friday. Weekends show lower total rides but longer ride duration, suggesting possible more leisure based trips compared to weekday commuting patterns.

```{r echo = FALSE}
# Heatmap: Hour × Day of week
hour_dow_summary <- citibike_nj_analysis %>%
  count(day_of_week, start_hour) %>%
  group_by(day_of_week) %>%
  mutate(percentage = n / max(n) * 100)

p_heatmap <- hour_dow_summary %>%
  ggplot(aes(x = start_hour, y = day_of_week, fill = percentage)) +
  geom_tile() +
  labs(
    title = "Heatmap: Ride Frequency by Hour and Day of Week in NJ City Area",
    x = "Hour of Day",
    y = "Day of Week",
    fill = "Relative\nFrequency (%)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  ) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  scale_x_continuous(breaks = seq(0, 23, by = 3))

print(p_heatmap)

```
The heatmap highlights strong weekday commuting patterns, with ride activity concentrated in the morning and especially in the late afternoon and early evening. Weekends show a more even distribution across midday hours, with fewer early-morning rides and a less pronounced afternoon peak.


# Analysis 3: Ride duration analysis in NJ city

```{r echo = FALSE, message = FALSE, warning = FALSE}
# Handle outliers using IQR method
calculate_outlier_bounds <- function(data, column) {
  q1 <- quantile(data[[column]], 0.25, na.rm = TRUE)
  q3 <- quantile(data[[column]], 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower_bound <- q1 - 1.5 * iqr
  upper_bound <- q3 + 1.5 * iqr
  return(list(lower = lower_bound, upper = upper_bound))
}

outlier_bounds <- calculate_outlier_bounds(citibike_nj_analysis, "ride_length_min")

# Create dataset without extreme outliers
duration_analysis <- citibike_nj_analysis %>%
  filter(ride_length_min <= outlier_bounds$upper)

# Duration statistics
duration_stats <- duration_analysis %>%
  summarise(
    n_rides = n(),
    mean_duration = mean(ride_length_min),
    median_duration = median(ride_length_min),
    sd_duration = sd(ride_length_min),
    q25 = quantile(ride_length_min, 0.25),
    q75 = quantile(ride_length_min, 0.75)
  )

kable(duration_stats, digits = 1)

# Plot: Duration distribution
p_duration_dist <- duration_analysis %>%
  ggplot(aes(x = ride_length_min)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  geom_vline(aes(xintercept = mean(ride_length_min)), 
             color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Distribution of Ride Durations in NJ City Area",
    subtitle = "Red line shows mean duration",
    x = "Ride Duration (minutes)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5)) +
  xlim(0, min(120, outlier_bounds$upper))

print(p_duration_dist)
```

Ride durations are short overall, clustering around 4–8 minutes, with a median of 6 minutes and a long tail of less frequent longer trips. The distribution is right-skewed, indicating that while most rides are brief, a smaller share extend well beyond the mean duration of about 7 minutes.

```{r echo = FALSE}
# Duration by member type
duration_member <- duration_analysis %>%
  group_by(member_type) %>%
  summarise(
    n_rides = n(),
    mean_duration = mean(ride_length_min),
    median_duration = median(ride_length_min),
    sd_duration = sd(ride_length_min),
    .groups = "drop"
  )

kable(duration_member, digits = 1)

# Plot: Duration by member type
p_duration_member <- duration_analysis %>%
  ggplot(aes(x = ride_length_min, fill = member_type)) +
  geom_density(alpha = 0.6) +
  labs(
    title = "Ride Duration Distribution by Member Type in NJ City Area",
    x = "Ride Duration (minutes)",
    y = "Density",
    fill = "Member Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "bottom"
  ) +
  scale_fill_manual(values = c("Annual Member" = "#00BFC4", "Casual Rider" = "#F8766D")) +
  xlim(0, min(120, outlier_bounds$upper))

print(p_duration_member)
```

Annual members tend to take shorter, more consistent rides, while casual riders have noticeably longer trip durations on average.

```{r echo = FALSE}
# Duration by time of day
duration_time <- duration_analysis %>%
  group_by(time_of_day) %>%
  summarise(
    mean_duration = mean(ride_length_min),
    n_rides = n(),
    .groups = "drop"
  )

kable(duration_time, digits = 1)

# Plot: Duration by time of day
p_duration_time <- duration_analysis %>%
  ggplot(aes(x = time_of_day, y = ride_length_min, fill = time_of_day)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Ride Duration by Time of Day in NJ City Area",
    x = "Time of Day",
    y = "Ride Duration (minutes)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  scale_fill_viridis_d(option = "C") +
  coord_cartesian(ylim = c(0, quantile(duration_analysis$ride_length_min, 0.95)))

print(p_duration_time)

```

Ride durations vary slightly by time of day. Mornings are the shortest on average, while afternoons, evenings, and nights are slightly longer and have more variable trips. The overall pattern suggests that rides later in the day tend to be more leisurely or flexible compared to morning travel.

# Analysis 4: User type comparisons in NJ city

```{r echo = FALSE}
# Overall distribution
member_dist <- citibike_nj_analysis %>%
  count(member_type) %>%
  mutate(percentage = n / sum(n) * 100)

kable(member_dist, digits = 1)

# Plot: Member distribution pie chart
p_member_pie <- member_dist %>%
  ggplot(aes(x = "", y = n, fill = member_type)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(
    title = "Overall Rider Type Distribution in NJ City Area",
    fill = "Rider Type"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "bottom"
  ) +
  scale_fill_manual(values = c("Annual Member" = "#00BFC4", "Casual Rider" = "#F8766D")) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5))

print(p_member_pie)
```

Annual members make up the majority while casual riders account for a smaller share of usage.

```{r echo = FALSE}
# Member patterns by day type
member_daytype <- citibike_nj_analysis %>%
  group_by(day_type, member_type) %>%
  summarise(
    n_rides = n(),
    avg_duration = mean(ride_length_min),
    .groups = "drop"
  ) %>%
  group_by(day_type) %>%
  mutate(
    total = sum(n_rides),
    percentage = n_rides / total * 100
  )

kable(member_daytype, digits = 1)

# Plot: Member patterns by day type
p_member_daytype <- member_daytype %>%
  ggplot(aes(x = day_type, y = percentage, fill = member_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Rider Type Distribution by Day Type in NJ City Area",
    x = "Day Type",
    y = "Percentage",
    fill = "Rider Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "bottom"
  ) +
  scale_fill_manual(values = c("Annual Member" = "#00BFC4", "Casual Rider" = "#F8766D")) +
  scale_y_continuous(labels = percent_format())

print(p_member_daytype)
```

```{r echo = FALSE}
# Member patterns by time of day
member_time <- citibike_nj_analysis %>%
  group_by(time_of_day, member_type) %>%
  summarise(
    n_rides = n(),
    .groups = "drop"
  ) %>%
  group_by(time_of_day) %>%
  mutate(
    total = sum(n_rides),
    percentage = n_rides / total * 100
  )

# Plot: Member patterns by time of day
p_member_time <- member_time %>%
  ggplot(aes(x = time_of_day, y = percentage, fill = member_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Rider Type Distribution by Time of Day in NJ City Area",
    x = "Time of Day",
    y = "Percentage",
    fill = "Rider Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  scale_fill_manual(values = c("Annual Member" = "#00BFC4", "Casual Rider" = "#F8766D")) +
  scale_y_continuous(labels = percent_format())

print(p_member_time)
```

The data shows that annual members make up the majority of weekday riding, accounting for over three-quarters of trips, while casual riders make up a larger share on weekends. By time of day, annual members remain the majority in every time period, specifically their share is highest in the morning when commuting is most common. However, annual member rides decline slightly later in the day as casual usage increases. Overall, the patterns reinforce that members drive weekday, commute oriented demand, while casual riders contribute more heavily during leisure times and weekends.

```{r echo = FALSE}
# Bike type preferences
bike_member <- citibike_nj_analysis %>%
  group_by(member_type, bike_type) %>%
  summarise(
    n_rides = n(),
    .groups = "drop"
  ) %>%
  group_by(member_type) %>%
  mutate(
    total = sum(n_rides),
    percentage = n_rides / total * 100
  )

kable(bike_member, digits = 1)

# Plot: Bike type preferences
p_bike_member <- bike_member %>%
  ggplot(aes(x = member_type, y = percentage, fill = bike_type)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Bike Type Preference by Rider Type in NJ City Area",
    x = "Rider Type",
    y = "Percentage",
    fill = "Bike Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "bottom"
  ) +
  scale_fill_viridis_d(option = "D") +
  scale_y_continuous(labels = percent_format())

print(p_bike_member)

```

The graph compares the distribution of bike types (classic, electric, and docked—used) by annual members compared to casual riders. Both groups mostly use classic bikes, but members use electric bikes slightly less often than casual riders. Casual riders also show a small amount of docked-bike usage, while members almost never use them.

